Climate indicators

### 2023-03-03 DRAFT_climate_indicators.Rmd
reg means

### Dummy data

@@

```{r}
reg <- sf::st_read("data/regions.shp")
#st_crs(reg)
```

@@

```{r}
temp_median_summer_ref_long <- temp_median_summer_ref %>% split("median_sd")
```

```{r}
temp_median_summer_ref_long <- setNames(temp_median_summer_ref_long, c("reference_upper", "sd"))
```

@@

```{r, eval=F, include=F}
temp_median_summer_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
```

@@

```{r}
# Mutate
# can I find a way to keep the v_ columns?

@@

@@

#### Plot time series

>>>>

#### Dummy data

@@

```{r}
reg <- sf::st_read("data/regions.shp", options = "ENCODING=UTF8")
#st_crs(reg)
```

Outline of Norway

```{r}
nor <- sf::st_read("data/outlineOfNorway_EPSG25833.shp")
```
Remoce marine areas from regions

```{r}
reg <- st_intersection(reg, nor)
```

```{r, fig.cap="Five accounting areas (regions)."}

tm_shape(reg) +
  tm_polygons(col="region")

@@

@@

```{r, eval=T, include=F}
temp_median_summer_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
```

```{r, eval=T}
temp_median_summer_ref_long <- temp_median_summer_ref %>% split("median_sd")
```

```{r, eval=T}
temp_median_summer_ref_long <- setNames(temp_median_summer_ref_long, c("reference_upper", "sd"))
```

@@

```{r, eval=F}
# Mutate
# can I find a way to keep the v_ columns?

@@

```{r, eval=F}
gc()
saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
```

```{r, include=FALSE}
y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
```

@@

The _real_ indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming. Therefore, the time series will be presented with yearly resolution, and the scaled, regional indicator values will be calculated for distinct time steps.

#### Plot time series as line graphs

We could also plot a time series as a series of maps, but then we need to avergae over this 5 time periods firs. We can get to that later.

To plot time series I first need to do a spatial aggregation. One option is to spatially aggregate the stars object directly, using the `reg` data set:

```{r, eval=F}
temp <- y_var_norm[1,,] %>%
  aggregate(reg, mean, na.rm=T)
plot(temp)
temp2 <- temp %>%
  st_as_sf() # too slow
```

However, this does not preserve the region name, and a subsequent st_intersection taks a little whie and gives some boundary problems.

I can also converting att pixels to points and then do the intersection.

```{r, eval=F}
temp_points <- y_var_norm[1,,] %>%
  st_as_sf(as_points = TRUE, merge = FALSE)
temp_points2 <- temp_points %>%
  st_intersection(reg) " too slow"
```

But this also took too long.

I can perhaps aggregate, and then turn to points and intersect, but that did not work perfect either.

I can rastereze the regions data set, turn it into a coded dimension and use st_apply, but thats tedious.

Let me try using `exactextract`, as a parallel to `aggregate`. This requires us to turn to the `terra` package for a bit.

```{r, eval=F}
(
regional_means <- rast(y_var_norm[,,]) %>%
  exact_extract(reg, fun = 'mean', append_cols = "region", progress=T) %>%
  setNames(c("region", names(y_var_norm)))
  )
```

That was fast!

```{r, eval=F}
saveRDS(regional_means, "temp/regional_means.rds")
```

Reshape and plot

```{r}
(
regional_means_long <- regional_means %>%
  as.data.frame() %>%
  rename(referenceUpper = reference_upper,
         referenceLow = reference_low,
         referenceLow2 = reference_low2,
         thresholdLow = threshold_low,
         thresholdHigh = threshold_high) %>%
  pivot_longer(!region) %>%
  separate(name, into=c("type", "year"), remove=F)
)
```

```{r}
temp3 <- temp2 %>%
  st_intersection(reg)
```
